---
title: "Lab 3 Homework"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the library
```{r}
library(tidyverse)
```

For this assignment we are going to work with a large dataset from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries. The data are messy, so for this assignment I am going to provide some help. The code I use will likely be useful in the future so keep it handy. First, load the data. **Read** the error message.  

```{r}
getwd
```

```{r}
fisheries <- readr::read_csv("~/Desktop/FRS_417/class_files-master/FAO_1950to2012_111914.csv")
## finds the fisheries data frame downloaded within the master class-files
```

## 1. Do you see any potential problems with the column names? Does the error message now make more sense?  
```{r}
names(fisheries)
## Column names are duplicated so R manipulated them to unique names by adding the "_1" after each duplicate column name
```

## 2. The `make.names()` command is helpful when there are issues with column names. Notice that although the names are still cumbersome, much of the problemtatic syntax is removed.
```{r}
names(fisheries) = make.names(names(fisheries), unique=T) #changes the duplicate column names to something else
names(fisheries)
```

## 3. Let's rename the columns. Use `rename()` to adjust the names as follows. Double check to make sure the rename worked correctly. Make sure to replace the old fisheries object with a new one so you can keep the column names.
```{r}
 ## rename this new name data frame as fisheries1
fisheries1 <- 
  fisheries %>% 
  rename(country     = Country..Country.,
         commname    = Species..ASFIS.species.,
         sciname     = Species..ASFIS.species._2,
         spcode      = Species..ASFIS.species._1,
         spgroup     = Species..ISSCAAP.group.,
         spgroupname = Species..ISSCAAP.group._1,
         region      = Fishing.area..FAO.major.fishing.area.,
         unit        = Measure..Measure.  )
##make sure to keep the period at the end of the column title
fisheries1
```

## 4. Are these data tidy? Why or why not, and, how do you know?
  ## The data is not tidy since there are multuple columns for each set of data. Typically, in tidy data, there should be one row and one column corresponding to one value.

##5. We need to tidy the data using `gather()`. The code below will not run because it is commented (#) out. I have added a bit of code that will prevent you from needing to type in each year from 1950-2012 but you need to complete the remainder `QQQ` and remove the `#`.
```{r}
fisheries_tidy <- 
  fisheries1 %>% 
  gather(num_range('X',1950:2012), key="year", value='catch')
##I wasn't sure where the year and catch come into the data frame, so I looked at the Github website under Professor Ledford's page.
## When looking at fisheries data frame, I saw that year and catch are part of the values within fisheries.
fisheries_tidy
```

6. Use `glimpse()` to look at the categories of the variables. Pay particular attention to `year` and `catch`. What do you notice?  
```{r}
glimpse(fisheries_tidy)
## I noticed how my values are entered as characters or dbl, and there are some values with an "X" in the value 
```

## 7. From question 6 you should see that there are a lot of entries that are missing. In R, these are referred to as NA's but they can be coded in different ways by the scientists in a given study. In order to make the data tidy, we need to deal with them. As a preview to our next lab, run the following code by removing the `#`. It removes the 'X' from the years and changes the `catch` column from a character into a numeric. This forces the blank entries to become NAs. The error "NAs introduced by coercion" indicates their replacement.
```{r}
fisheries_tidy <- 
  fisheries_tidy %>% 
   mutate(
     year= as.numeric(str_replace(year, 'X', '')), #replaces in X in the key "year" with a blank space instead
     catch= as.numeric(str_replace(catch, c(' F','...','-'), replacement = '')) #replace character strings with NA
     )
fisheries_tidy
```

8. Are the data tidy? Why?  
## The data are now tidy because each row corresponds to row of data

## 9. You are a fisheries scientist studying cephalopod catch during 2008-2012. Identify the top five consumers (by country) of cephalopods (don't worry about species for now). Restrict the data frame only to our variables of interest.
```{r}
fisheries_tidy %>%
  select(country, spgroupname, year, catch) %>% 
  filter(
    year==2012 | year==2011 | year==2010 | year==2009 | year==2008,
    spgroupname=="Squids, cuttlefishes, octopuses") %>%
  ## A cephalopod is any member of the molluscan class Cephalopoda such as a squid, octopus or nautilus.
  arrange(desc(catch)) ##arranges in descending order of number of catches
```
  
10. Let's be more specific. Who consumes the most `Common cuttlefish`? Store this as a new object `cuttle`.
```{r}
cuttle<- fisheries_tidy %>% 
  select(country, commname, spgroupname, year, catch) %>% 
  filter(year==2012 | year==2011 | year==2010 | year==2009 | year==2008,
         commname=="Common cuttlefish") %>% 
  arrange(desc(catch))
cuttle
##arranges all the Common Cuttlefish by country of decreased catch order
```

