---
title: "Lab 6 Homework"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(skimr)
library("RColorBrewer")
```

## Gapminder
For this assignment, we are going to use the dataset [gapminder](https://cran.r-project.org/web/packages/gapminder/index.html). Gapminder includes information about economics, population, and life expectancy from countries all over the world. You will need to install it before use.
```{r}
##install.packages("gapminder")
```

```{r}
library("gapminder")
```

Please load the data into a new object called gapminder.
```{r}
gapminder <- 
  gapminder::gapminder
gapminder
```

1. Explore the data using the various function you have learned. Is it tidy, are there any NA's, what are its dimensions, what are the column names, etc.
## The data is tidy because each different observation of the variables are in a different row. In addition, the data is easy to read because each of the 6 variables corresponds to one column
```{r}
colnames(gapminder) ##shows column names of gapminder
```
```{r}
glimpse(gapminder) 
##1,704 observations with 6 variables (columns)
##Dimensions: 1,704 x 6 (rows x columns)
```
```{r}
gapminder<-
 gapminder %>% 
  na_if(-999)
##Replaces all -999 values with NA so that it is easier to track where there is an NA with the next line of code
```

```{r}
number_nas= sum(is.na(gapminder))
gapminder %>%
  purrr::map_df(~ sum(is.na(.))) %>%  ##shows how many NA values and where they are within the data frame
  tidyr::gather(variables, number_nas) %>% ##I had to get help from GitHub to understand the last 2 lines here
  arrange(desc(number_nas))
## This shows that there are no NA values in gapminder
```

2. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer on average. Make a quick plot below to visualize this relationship.
```{r}
gapminder %>% 
  ggplot(aes(x=gdpPercap, y=lifeExp))+
  geom_point()+
  labs(title = "Life Expectancy vs GDP",
       x = "Per Capita GDP",
       y= "Life Expectancy")
##There seems to be a positive relationship between life expectancy and gdp. The longer a person lives, the higher their per capita GDP
```

3. There is extreme disparity in per capita GDP. Rescale the x axis to make this easier to interpret. How would you characterize the relationship?
```{r}
gapminder %>% 
  ggplot(aes(x=log10(gdpPercap), y=lifeExp))+
  geom_point()+
   labs(title = "Life Expectancy vs GDP",
       x = "Per Capita GDP",
       y= "Life Expectancy")
##There seems to be a positive relationship between life expectancy and gdp. The longer a person lives, the higher their per capita GDP
```

4. This should look pretty dense to you with significant overplotting. Try using a faceting approach to break this relationship down by year.
```{r}
gapminder %>% 
  ggplot(aes(x = log10(gdpPercap), y=lifeExp)) +
  geom_point() +
   labs(title = "Life Expectancy vs GDP",
       x = "Per Capita GDP",
       y= "Life Expectancy")+
  facet_wrap(~year) ##"facet wrap by tilda taxon"- provides individual histograns for each of the taxa
##Shows the positive linear relationship between life expectancy and gdp by year
```

5. Simplify the comparison by comparing only 1952 and 2007. Can you come to any conclusions?
```{r}
gapminder %>% 
  filter(year==1952 | year==2007) %>%
  ggplot(aes(x=log10(gdpPercap), y=lifeExp))+
  geom_point(color="orchid")+
  facet_wrap(~year)+
  theme_light()
```

6. Let's stick with the 1952 and 2007 comparison but make some aesthetic adjustments. First try to color by continent and adjust the size of the points by population. Add `+ scale_size(range = c(0.1, 10), guide = "none")` as a layer to clean things up a bit.
```{r}
gapminder %>%
  filter(year==1952 | year==2007) %>% 
  ggplot(aes(x=log10(gdpPercap), y=lifeExp, color=continent, size=pop))+
  geom_point()+
  facet_grid(~year)+
  scale_size(range = c(0.1, 10), guide = "none")
```


7. Although we did not introduce them in lab, ggplot has a number of built-in themes that make things easier. I like the light theme for these data, but you can see lots of options. Apply one of these to your plot above.
```{r}
?theme_light
```

8. What is the population for all countries on the Asian continent in 2007? Show this as a barplot.
```{r}
gapminder %>% 
  filter(continent=="Asia", year=="2007") %>% 
  ggplot(aes(x=country, y=pop, fill=country))+
  geom_bar(stat="identity")+
   labs(title = "China Population over the Years",
       x = "Popualtion",
       y= "Country")+
  coord_flip()
```

9. You should see that China's population is the largest with India a close second. Let's focus on China only. Make a plot that shows how population has changed over the years.
```{r}
gapminder %>% 
  filter(country=="China") %>% 
   ggplot(aes(x=year, y=pop))+
  geom_bar(stat = "identity")+
  labs(title = "China Population over the Years",
       x = "Year",
       y= "Population")+
  ylim(0, 1.5E+09)
## The barplot shows the continuous increase in the China Population over the years; Increasing trendline
```


10. Let's compare China and India. Make a barplot that shows population growth by year using `position=dodge`. Apply a custom color theme using RColorBrewer.
```{r}
gapminder %>% 
  filter(country=="China" | country=="India") %>% 
  ggplot(aes(x=year, y=pop, fill=country))+
    geom_bar(stat="identity", position="dodge")+
  scale_fill_brewer(palette = "Dark4")
```