---
title: "Visualizing Data 3"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

```{r message=FALSE, warning=FALSE}
homerange <- 
  readr::read_csv("~/Desktop/FRS_417/class_files-master/data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", " ", "NA", "#N/A", "-999", "\\")) ##Replaces the -999 values w NA immediately as the data file is read into R
homerange
```
```{r}
names(homerange)
```


## Barplots revisited
At this point you should be able to build a barplot that shows counts of observations in a variable using `geom_bar()`. But, you should also be able to use `stat="identity"` to specify both x and y axes.  

## Here is the plot using `geom_bar(stat="identity")`
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% #filter for salmonid fish only
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_bar(stat="identity")
```

## Practice
Although we did not use it last time, `geom_col()` is the same as specifying `stat="identity"` using `geom_bar()`.     
1. Make the same plot above, but use `geom_col()`
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% #filter for salmonid fish only
  ggplot(aes(x=common.name, y=log10.mass, fill=common.name))+ ##fills color based on the common names
  geom_col()
```

## Barplots and multiple variables
Last time we explored the `fill` option in boxplots as a way to bring color to the plot; i.e. we filled by the same variable that we were plotting. What happens when we fill by a different categorical variable? 
Let's start by counting how many obervations we have in each taxonomic group.
```{r}
homerange %>% 
  count(taxon)
```

Now let's make a barplot of these data.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=taxon))+ ## fills by taxon
  geom_bar(alpha=0.9, na.rm=TRUE)+
  coord_flip()+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group")
```

By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+ ##makes a stacked bar blot with trophic guild and taxon
  geom_bar(alpha=0.9, na.rm=T)+
  coord_flip()+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")
```

We can also have the proportions of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+ ##filling by a different categorical variable than x axis (taxon)
  geom_bar(positiion="dodge")+
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")+
coord_flip()
```

## Practice
1. Make a barplot that shows locomotion type by `primarymethod`. Try both a stacked barplot and `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x=locomotion, fill=primarymethod))+
  geom_bar(position="dodge") ##without position="dodge", the bar plot would be stacked (may be harder to read)!
```

```{r}
homerange %>% 
  ggplot(aes(x=locomotion, fill=primarymethod)) +
  geom_bar(position=position_fill(), stat="count")+ ## stat="count" is default but we are adding it in for reference purposes
  scale_y_continuous(labels=scales::percent) ##changes scales to percent format
```

## Histograms and Density Plots

## Need to specify number of counts and bends in a histogram, but this may not always be clear based on types of data given
## Histograms are frequency distribution plots
What does the distribution of body mass look like in the homerange data?
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(fill="steelblue", alpha=0.8, color="black")+ ##using geom_histogram and making it "pretty"; alpha control the transparency levels; color describes the outline color
  labs(title = "Distribution of Body Mass")
```

```{r}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram()+ ##using geom_histogram
  labs(title = "Distribution of Body Mass")
## This is the histogram without making it more pretty!
```


`Density plots` are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.; ## No bins in the histogram; graph will look like a curve
```{r}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_density(fill="steelblue", alpha=0.4)+
  labs(title = "Distribution of Body Mass")
```

##Both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = ..density..), binwidth = .4, fill="steelblue", alpha=0.8, color="black")+
  geom_density(color="red")+
  labs(title = "Distribution of Body Mass")
##This will plot both the density graph and the histogram plot 
```

## Practice
1. Make a histogram of `log10.hra`. Make sure to add a title.
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra))+
  geom_histogram(fill="yellow", alpha=0.3, color="black")+
  labs(title = "Distribution of HRA")
```


2. Now plot the same variable using `geom_density()`.
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra))+
  geom_density(fill="orange", alpha=0.3, color="black")+
  labs(title = "Distribution of HRA")
```

3. Combine them both!
```{r}
homerange %>% 
  ggplot(aes(x=log10.hra))+
  geom_histogram(aes(y=..density..), binwidth = 0.5, fill="orchid", alpha=0.7, color="black")+
  geom_density(color="brown")
  labs(title = "Distribution of HRA")
```
