---
title: "Lab 4 Homework"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
---
## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Aren't mammals fun? Let's load up some more mammal data. This will be life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*  

```{r}
life_history <- readr::read_csv("~/Desktop/FRS_417/class_files-master/mammal_lifehistories_v2.csv")
```

```{r}
names(life_history)
```

Rename some of the variables. Notice that I am replacing the old `life_history` data.
```{r}
life_history <- 
  life_history %>% 
  dplyr::rename(
          genus        = Genus,
          wean_mass    = `wean mass`,
          max_life     = `max. life`,
          litter_size  = `litter size`,
          litters_yr   = `litters/year`
          )
life_history
```

## 1. Explore the data using the method that you prefer. Below, I am using a new package called `skimr`. If you want to use this, make sure that it is installed.
```{r}
install.packages("skimr")
##installs a new package (not the tidyverse!)
```

```{r}
library("skimr")
```

```{r}
life_history %>% 
  skimr::skim()
##skimr is like another summary function
```

## 2. Run the code below. Are there any NA's in the data? Does this seem likely?
```{r}
msleep %>% 
  summarize(number_nas= sum(is.na(life_history)))
## Technically, there are no NA's, but it is possible that there is another value such as -999 that represents a null value
```

## 3. Are there any missing data (NA's) represented by different values? How much and where? In which variables do we have the most missing data? Can you think of a reason why so much data are missing in this variable?
```{r}
life_history<-
  life_history %>% 
  na_if(-999)
## In class, we learned how -999 is sometimes used to represent values where there is no data entry, so I checked for any -999 values that may be in place for NA. If there was an -999, I replacd it with an NA so that it would be easier to identify where the null value is located within the data frame. 
```

```{r}
number_nas= sum(is.na(life_history))
life_history %>%
  purrr::map_df(~ sum(is.na(.))) %>%  ##shows how many NA values and where they are within the data frame
  tidyr::gather(variables, number_nas) %>% ##I had to get help from GitHub to understand the last 2 lines here
  arrange(desc(number_nas))
##There might be many missing wean and newborn values missing since not all are measured exactly at birth
```

## 4. Compared to the `msleep` data, we have better representation among taxa. Produce a summary that shows the number of observations by taxonomic order.
```{r}
names(life_history) ##I'm using this to get a general idea of the column headers so I know which notation to use within my code
```

```{r}
life_history %>% 
  group_by(order) %>% ##groups by taxonomic order
  summarize(n())   ## n represents the number of observations; each order has a certain number of observations
```

## 5. Mammals have a range of life histories, including lifespan. Produce a summary of lifespan in years by order. Be sure to include the minimum, maximum, mean, standard deviation, and total n.
```{r}
life_history %>%
  mutate(lifespan=max_life/12) %>% ##This changes lifespan values from months into years  
  group_by(order) %>%
  summarize(min_lifespan=min(lifespan, na.rm=TRUE), ##return lifespan values for which the values are not NA
            max_lifespan=max(lifespan, na.rm=TRUE),
            mean_lifespan=mean(lifespan, na.rm=TRUE),
            sd_lifespan=sd(lifespan, na.rm=TRUE),
            total_lifespan=n()
  )
```

## 6. Let's look closely at gestation and newborns. Summarize the mean gestation, newborn mass, and weaning mass by order. Add a new column that shows mean gestation in years and sort in descending order. Which group has the longest mean gestation? What is the common name for these mammals?
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_gestation=mean(gestation, na.rm=TRUE), ##summarizes mean for specified values
            mean_newborn_mass=mean(newborn, na.rm=TRUE),
            mean_wean_mass=mean(wean_mass, na.rm=TRUE)) %>% 
  arrange(desc(mean_gestation)) 
```
## Proboscidea has longest mean gestation; "The Proboscidea (from the Greek προβοσκίς and the Latin proboscis) are a taxonomic order of afrotherian mammals containing one living family (Elephantidae) and several extinct families"