---
title: "Tidy Data 2 and Summarize"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
## tidy data: every column is essentially its own vector; there is one data type; names of columns are of the same class (ex. characters rather than data)

## Load the tidyverse
```{r}
library("tidyverse")
```

## Tidyr
+ `gather()` and `spread()` convert data between wide and long format  
+ `separate()` and `unite()` separates or unites information in columns 

## gather()
Recall that we use `gather()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
?USArrests
## pulls up a brief article about this data set
```

```{r}
arrests <- 
  USArrests %>% 
  mutate(State=rownames(USArrests)) %>% 
  select(State, Murder, Assault, Rape)
arrests ##shows arrests per 100,000 individuals per state
```

## Practice
1. Are these data tidy? Please use `gather()` to tidy the data.
## Data is not tidy because there is not one row that corresponds to the 3 sets of numerical data; the state has 3 columns representing affiliated with it
```{r}
tidy_arrests<-
  arrests %>% 
  gather(Murder, Assault, Rape, key="Crime", value="per100k")
tidy_arrests
```

2. Restrict the data to assualt only. Sort in ascending order.
```{r}
##filter the tidy_arrests since it is already gathered; use the filter command to find assaults
tidy_arrests %>% 
  filter(Crime=="Assault") %>% 
  arrange(desc(per100k))
```


## spread()
The opposite of `gather()`. You use `spread()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.
```{r}
country <- c("Afghanistan", "Afghanistan", "Afghanistan", "Afghanistan", "Brazil", "Brazil", "Brazil", "Brazil", "China", "China", "China", "China")
year <- c("1999", "1999", "2000", "2000", "1999", "1999", "2000", "2000", "1999", "1999", "2000", "2000")
key <- c("cases", "population", "cases", "population", "cases", "population", "cases", "population", "cases", "population", "cases", "population")
value <- c(745, 19987071, 2666, 20595360, 37737, 172006362, 80488, 174504898, 212258, 1272915272, 213766, 1280428583)

tb_data <- data.frame(country=country, year=year, key=key, value=value)
tb_data ##shows tuberculosis infection rates in different countries by year
```

When using `spread()` the `key` is the variable that you are spreading.
```{r}
tb_data %>% 
  spread(key=key, value=value) ##makes data that are tidy; each column represents it own vector
##key takes the number of categories and creates that same number of colums (key contains cases and population, so these will become 2 columns after using the spread() function)
```

## Practice
1. Run the following to build the `gene_exp` data frame.
```{r}
id <- c("gene1", "gene1", "gene2", "gene2", "gene3", "gene3")
type <- c("treatment", "control", "treatment", "control","treatment", "control")
L4_values <- rnorm(6, mean = 20, sd = 3)
```

```{r}
gene_exp <- 
  data.frame(gene_id=id, type=type, L4_values=L4_values)
gene_exp
```

2. Are these data tidy? Please use `spread()` to tidy the data.
```{r}
gene_exp %>% 
  spread(type, L4_values) ##could also write as spread(key=type, value=L4_values)
##separates type into control and treatment columns
```


## summarize()
summarize() will produce summary statistics for a given variable in a data frame. For example, in homework 2 you were asked to calculate the mean of the sleep total column for large and small mammals. We did this using a combination of tidyverse and base R commands, which isn't very efficient or clean. It also took two steps.
```{r}
?msleep ##produces information about the msleep data frame
##name of data frame msleep is "mammals sleep"
```

From homework 2.
```{r}
large <- msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt>=200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total) ##find the means of the bodyweight since we are looking at "large" data frame which is filtered for bodyweight
```

```{r}
names(msleep)
```

We can accomplish the same task using the `summarize()` function in the tidyverse and make things cleaner.
```{r}
msleep %>% 
  filter(bodywt>=200) %>%
  summarize(mean_sleep_lg=mean(sleep_total)) ##makes a new column that is the mean of the sleep total
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
    filter(bodywt>=200) %>% 
    summarize(mean_sleep_lg = mean(sleep_total), 
              min_sleep_lg = min(sleep_total),
              max_sleep_lg = max(sleep_total),
              sd_sleep_lg=sd(sleep_total), ##gives the standard deviation for total sleep (compares values to the mean total sleep)
              total = n()) ## returns the length of the vector
##shows the summary of the mean of the requested variables
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of vector), first() (returns first value in vector), last() (returns last value in vector) and n_distinct() (number of distinct values in vector).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>% 
  summarize (ngenera=n_distinct(genus)) ##looking at number of genera within the genus column
```

2. What are the min, max, and mean body weight for all of the mammals? Be sure to include the total n.
```{r}
msleep %>% 
  summarize(minbodywt=min(bodywt),
            maxbodywt=max(bodywt),
            meanbodywt=mean(bodywt),
            total=n()) ##returns the total number of observations within that column
```

## show how many types of "vores" there are within the msleep data frame
```{r}
msleep %>% 
  count(vore)
##vore is a column
```

##group_by()
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the dataset is helpful, what if we were interested in body weight by feeding ecology?
```{r}
msleep %>%
  group_by(vore) %>% ##vore=trophic types; herbivore, omnivore, carnivore, etc; we are grouping by feeding ecology
  summarize(min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            mean_bodywt=mean(bodywt),
            total=n()) #gives number of observations within each set
## in a huge data set, you can now group by however you want to within a data set; specify which data you are looking at
## use in data sets that you don't know very well; statistical summary of basic info provided within the frame
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>% ##groups by taxonomic order
  summarize(mean_brain_wt=(mean(brainwt)))
##There are many different Artiodactyla; if there is even 1 Ariodactyla without the brain data entered, it will return NA for all the Artiodactyla!
```


2. What does `NA` mean?
##some values return NA bc values are not present for that order

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea?
```{r}
msleep %>% 
  group_by(order) %>%
  summarize(mean_brain_wt=mean(brainwt,na.rm=TRUE)) ##return brain weight for which the values are not NA; Cratacea has a problem since it does not contain any brain weight values
```

```{r}
msleep %>% 
  filter(order=="Cetacea")
##there is no brain weight for Cetacea; can't calculate the mean for something that is not there
```
