---
title: "Data Frames and filter()"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
##Practice and Review from previous class
#vectors are strings of the same type
```{r}
vector_1 <- c(1, 2, 3, "orange", "blue")
vector_1
```
## The vector will print as a string without any problem; however, there is a problem when we try to identify what class the vector is;
```{r}
class(vector_1)
```

#This is the beginning of Lab2_2
```{r}
library("tidyverse")
```
```{r}
getwd()
## This will show where our R Markdown is being saved
```
```{r}
fish <- readr::read_csv("~/Desktop/FRS_417/class_files-master/Gaeta_etal_CLC_data.csv")
## Open a file from tidyverse that contains data about fish scales
## Use read_csv since we are using tidyverse
## readr: use readr to read comma separated values (CSV) located in the specified directory where class file master is located
```
## To look at the actual file, go to environment --> fish--> click on the spreadsheet button on the right side. This should pull up a spreadsheet w all the data values

## Summary functions
Once a data frame has been imported, you may want to get an idea of its contents and dimensions. This may seem self-evident for small files but this can be a challenge for large data sets. Summary functions help us get a general idea of the contents of a data frame.
```{r}
names(fish)
```

```{r}
head(fish)
```
## dbl = double precision numberic

## str = structure
```{r}
str (fish) # can use glimpse instead of str
```

```{r}
summary (fish) ## gives information about each column within the vector
## gives statistics for each column! VERY HELPFUL
```

```{r}
glimpse (fish) #gives observations, number of variables, and class each value belongs to
```


```{r}
nrow(fish) #the number of rows
```

```{r}
ncol(fish) #the number of columns
```

```{r}
dim(fish) #total dimensions
## prints out 
```
## there are 4033 objects of 6 variables

```{r}
colnames(fish) #column names
```

## Practice
## 1. Load the data `mammal_lifehistories_v2.csv` and place it into a new object called `mammals`.
```{r}
mammals <- readr::read_csv("~/Desktop/FRS_417/class_files-master/mammal_lifehistories_v2.csv")
## pulls the class file from the class master we downloaded from GitHub
```

## 2. Provide the dimensions of the data frame.
```{r}
dim(mammals)
## shows that the data frame has 1440 rows and 13 columns
```

## 3. Display the column names in the data frame. 
```{r}
colnames(mammals) 
##displays the 13 column names
```

## 4. Use str() to show the structure of the data frame and its individual columns; compare this to glimpse()
```{r}
str(mammals) ## shows the structure of the data frame
```
```{r}
glimpse(mammals) ## shows glimpse of the data frame
```

## 5. Print out the first few rows of the data using the function head(). 
```{r}
head(mammals) 
## prints the first 6 rows and the first 10 columns
```

## Dplyr: filter()
A core package in the **tidyverse** is `dplyr`. This package allows you to transform your data in many different ways to focus on variables of interest. This helps keep your data clean, easier to work with, and easier for other people to understand.  
## Will allow us to extract info from the data frame
```{r}
names (fish)
```

## If we are only interested in fish that come from Alabama lakes, we would use the filter()
```{r}
filter (fish, lakeid == "AL") ##lakeid represents the lake (state) where the fish is found
## use "==" to indicate which value we are looking for
```

## filter fish frame for scale length greater than or equal to 350
```{r}
filter (fish, length >=350)
```

##find fish with length = 350 in Florida lakes
```{r}
filter(fish, length ==350 & lakeid =="FD")
## nothing is returned since there are no lakes in Florida with length of 350
```

## find fish with length 167 and length 175
```{r}
filter (fish, length ==167 & length ==175)
## nothing returns since no fish can have both lengths; we should use OR (|) command instead
```

## find fish with either length 167 or length 175
```{r}
filter (fish, length ==167 | length ==175)
```

##"!" means not. Filter fish for lengths that are not 175
```{r}
filter (fish, !length==175)
```

## Practice
## 1. Filter the `fish` data to include the samples from lake `DY`
```{r}
filter(fish, lakeid=="DY")
```

## 2. Filter the data to include all lakes except AL.
```{r}
filter(fish, !lakeid=="AL")
## returns data frame for all fish not from Alabama
```

##3. Filter the data to include all lakes except AL and DY.
```{r}
filter (fish, !lakeid=="AL" & !lakeid=="DY")
## won't print fish data if it comes from AL or from DY
```

##4. Filter the data to include all fish with a scale length greater than or equal to 11.
```{r}
filter(fish, length >=11)
```

##5. Filter the data to include fish only from lake AL and with a scalelength greater than or equal to 2 and less than or equal to 4.
```{r}
filter (fish, lakeid=="AL" & (length>=2 & length<=4))
## no rows outputted since there are no fish with the given parameters
```

