---
title: "Data Visualization 1"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
##Plots are built in layers
## What makes a good chart? Its elegance and simplicity. It provides a clean, clear visual of the data without being overwhelming to the reader.
## Data Types
+  `discrete` quantitative data that only contains integers
+ `continuous` quantitative data that can take any numerical value
+ `categorical` qualitative data that can take on a limited number of values

## Basics: **plot= data + geom_ + aesthetics**  
## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("skimr")
```

```{r}
?iris ##provides overview of iris data
```

```{r}
?iris
names(iris) ##gives category (column names) for iris
```

```{r}
iris <-
  iris ##puts iris data into a new object so that we can look at iris data quickly
```

## To make a plot, we need to first specify the data and map the aesthetics. The aesthetics include how each variable in our dataset will be used. In the example below, I am using the aes() function to identify the x and y variables in the plot.
```{r}
ggplot(data=iris, mapping=aes(x=Species, y=Petal.Length)) ## we named Species and Petal.length based on names(iris)
```

## Notice that we have a nice background, labeled axes, and even values of our variables- but no plot. This is because we need to tell ggplot what type of plot we want to make. This is called the geometry or `geom()`.Here we specify that we want a boxplot, indicated by `geom_boxplot()`.
```{r}
ggplot(data=iris, mapping=aes(x=Species, y=Petal.Length))+
  geom_boxplot() ##creates a boxplot based on the ggplot data from x and y axes
## We add a "+" bc plots are built in layers
## First layer is the data in the plot; Second layer is the design of the plot
```

## Practice
1. Take a moment to practice. Use the iris data to build a scatterplot that compares sepal length vs. sepal width. Use the cheatsheet to find the correct `geom_` for a scatterplot.
```{r}
ggplot(data=iris, mapping=aes(x=Sepal.Width, y=Sepal.Length))+
  geom_point() ##plots data in iris (Species.Width, Sepal.Length) as a scatterplot diagram
```

 ##Two standard plots: 1) scatterplots and 2) barplots.


**Database of vertebrate home range sizes.Import from classfiles**

```{r message=FALSE, warning=FALSE}
homerange <- readr::read_csv("~/Desktop/FRS_417/class_files-master/Tamburelloetal_HomeRangeDatabase.csv")
```
```{r}
homerange
```

A little bit of cleaning to focus on the variables of interest. Good `dplyr` practice!
```{r}
homerange <- ##modifies homerange to select for these categories
  homerange %>%  
  select(common.name, taxon, family, genus, species, mean.mass.g, log10.mass, mean.hra.m2, log10.hra, preymass, log10.preymass, trophic.guild)
```

```{r}
names(homerange)
```

### 1. Scatter Plots
##Scatter plots are good at revealing relationships that are not readily visible in the raw data. For now, we will not add regression lines or calculate any r^2^ values. In this case, we are exploring whether or not there is a relationship between animal mass and homerange. We are using log transformed values because there is a very large difference in mass and homerange among the different  species in the data.

##Scatter plot deals with numerical data, not categorical
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_point()
## Is there an association between size of the beast and the area it requires to live? Hopefully, the answer is yes since larger beasts require more space to live

## The scatter plot is linear--> there is a correlation between mass and homerange area
## There is a lot of overlap between some points (dense areas) --> we can use geom_jitter to clean up the plot
```

## In big data sets with lots of similar values, overplotting can be an issue. `geom_jitter()` is similar to `geom_point()` but it helps with overplotting by adding some random noise to the data and separating some of the individual points.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_jitter()
```

##Adding in a linear regression line
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra)) +
  geom_jitter()+
  geom_smooth(method=lm, se=FALSE) 
## method = lm means that we are adding in a smooth lm (linear model)
```

### 2A. Bar Plot: `stat="count"` When making a bar graph, the default is to count the number of observations in the specified column. This is best for categorical data and is defined by the aesthetic `stat="count"`. Here, I want to know how many carnivores vs. herbivores are in the data.  
## Bar plots are best suited for categorical data

## Notice that we can use pipes and the `mapping=` function is implied by `aes` and so is often left out (Bc ggplot already understands it's mapping)  

#### we cannot specify a y variable for a geom_bar. By default, it creates a count on the y-axis
```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild))+ ##trophic guild represents carnivore, herbivore, omnivore, etc
  geom_bar(stat="count") # I am identifying stat="count" but this isn't necessary since it is default
## x axis is the trophic guild and y axis represents the number of each type of trophic guild (the count of each)
```
##we can also remove the stat="count" since the y-axis will automatically create a count-like axis
```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild))+ ##trophic guild represents carnivore, herbivore, omnivore, etc
  geom_bar()
```


### 2B. Bar Plot: `stat="identity"`
`stat="identity"` allows us to map a variable to the y axis so that we aren't restricted to count. Let's use our dplyr skills to first filter out carnivorous mammals and see which families have the highest mean body weight.
```{r}
carni_mammals <- 
  homerange %>% 
  filter(taxon=="mammals", 
         trophic.guild=="carnivore") %>% 
  group_by(family) %>% ## groups by families that are within the carnivourous trophic guild
  summarize(count=n(),
            mean_body_wt=mean(log10.mass)) %>% 
  arrange(desc(mean_body_wt))
carni_mammals ## shows the carnivores and their families by decreasing body weight
```

##Visualize the previous code(body weight of carnivores in different families) by creating a plot
Now let's plot the data using `stat="identity"` to help visualize these differences.
```{r}
carni_mammals%>% 
  ggplot(aes(x=family, y=mean_body_wt))+
  geom_bar(stat="identity")
##plots the different body weights on the y-axis instead of count
```

## This looks nice, but the x axis is a mess and needs adjustment. We make these adjustments using `theme()`. 
## Adjust the angle of the text on the x-axis
```{r}
carni_mammals%>% 
  ggplot(aes(x=family, y=mean_body_wt))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 60, hjust=1)) ##Change the angle of the text on the x-axis to 60 degrees
## hjudt=1 adjusts the length of the text next to the bar (placement of the text on the x-axis)
```

##We can make this a little bit better by reordering.
```{r}
carni_mammals%>% 
  ggplot(aes(x=reorder(family, -mean_body_wt), y=mean_body_wt))+ ##reorder will label data as ascending or descending based on if mean_body_wt is - or +
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```

##Turn the plot on its side; flip the coordinates.
```{r}
carni_mammals%>% 
  ggplot(aes(x=reorder(family, -mean_body_wt), y=mean_body_wt))+
  geom_bar(stat="identity")+
  coord_flip()
```

## Practice
Filter the `homerange` data to include `mammals` only.
```{r}
mammals <- 
  homerange %>% 
  filter(taxon=="mammals") %>% 
  select(common.name, family, genus, species, trophic.guild, mean.mass.g, log10.mass, mean.hra.m2, log10.hra, preymass, log10.preymass)
```

```{r}
mammals
```

##1. Are there more herbivores or carnivores in mammals? Make a bar plot that shows their relative proportions.
```{r}
mammals %>% 
  ggplot(aes(x=trophic.guild))+ ##trophic guild represents carnivore, herbivore, omnivore, etc
  geom_bar()
## We can tell that there are more herbivores by looking at the bar plot of mammals and plotting the trophic guild on the x axis
```

```{r}
names(mammals)
```

##2. Is there a positive or negative relationship between mass and homerange? How does this compare between herbivores and carnivores? Make two plots that illustrate these eamples below.  
## USe dyplr functions such as filter, select, etc
carnivores
```{r}
mammals %>% 
  filter (trophic.guild=="carnivore") %>%
  ggplot(aes(x=log10.mass, y=log10.hra)) + ##make a plot based off of the carnivore trophic guild 
  geom_point()+
  geom_smooth(method=lm, se=FALSE) 
```

herbivores
```{r}
mammals %>% 
  filter (trophic.guild=="herbivore") %>%
  ggplot(aes(x=log10.mass, y=log10.hra)) + ##make a plot based off of the herbivore trophic guild 
  geom_point()+
  geom_smooth(method=lm, se=FALSE) 
```

##3. Make a barplot that shows the masses of the top 10 smallest mammals in terms of mass. Be sure to use `stat'="identity"`.
## Hard Way
```{r}
homerange %>% 
  filter(taxon=="mammals") %>% 
  select(common.name, mean.mass.g) %>% 
  arrange (mean.mass.g)
## cinereus shrew is the mammal w smallest mass 
```

## Easy way
```{r}
mammals %>% 
  top_n(-10, log10.mass) %>% ##gives the top 10 smallest mammals
  ggplot(aes(x=common.name, y = log10.mass))+
  geom_bar (stat="identity")+
  coord_flip()
```

## Make the bar plot from #3 look prettier! This will show the bar plot in increasing order!
```{r}
 mammals %>% 
  top_n(-10, log10.mass) %>% ##gives the top 10 smallest mammals
  ggplot(aes(x=reorder(common.name,log10.mass), y = log10.mass))+
  geom_bar (stat="identity")+
  coord_flip()
```
