---
title: "Lab 5 Homework"
author: "Pallavi Malladi"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Mammals Life History
Let's revisit the mammal life history data to practice our `ggplot` skills. Some of the tidy steps will be a repeat from the homework, but it is good practice. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*

## 1. Load the data.
```{r}
life_history <- readr::read_csv("~/Desktop/FRS_417/class_files-master/mammal_lifehistories_v2.csv")
```


## 2. Use your preferred function to have a look. Do you notice any problems?
```{r}
life_history
## Looking at the data, we see that the data in the frame is currently tidy. The problem in life_history is that there were a relative large number of null values (represented by -999). We replaced -999 with NA values.
```

## 3. There are NA's. How are you going to deal with them?
## The NA's are represented by -999
```{r}
life_history<-
  life_history %>% 
  na_if(-999)
## In class, we learned how -999 is sometimes used to represent values where there is no data entry, so I checked for any -999 values that may be in place for NA. If there was an -999, I replacd it with an NA so that it would be easier to identify where the null value is located within the data frame. 
```

## 4. Where are the NA's? This is important to keep in mind as you build plots.
```{r}
number_nas= sum(is.na(life_history))
life_history %>%
  purrr::map_df(~ sum(is.na(.))) %>%  ##shows how many NA values and where they are within the data frame
  tidyr::gather(variables, number_nas) %>% ##I had to get help from GitHub to understand the last 2 lines here
  arrange(desc(number_nas))
```

## 5. Some of the variable names will be problematic. Let's rename them here as a final tidy step.
```{r}
life_history <-
  life_history %>% 
  dplyr::rename(
          genus        = Genus,
          wean_mass    = `wean mass`,
          max_life     = `max. life`,
          litter_size  = `litter size`,
          litters_yr   = `litters/year`
          )
life_history
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## 6. What is the relationship between newborn body mass and gestation? Make a scatterplot that shows this relationship. 
```{r}
names(life_history) ##shows column names for data frame so we know what to use for axes headers
```

```{r}
ggplot(data=life_history, mapping=aes(x=gestation, y=log10(newborn))) + ##wean mass=newborn body mass
  geom_jitter() +
   geom_smooth(method=lm, se=FALSE) ## method = lm means that we are adding in a smooth lm (linear model)
##I took the log10 of wean mass because the y-axis had a range that was too large for the graph. With log, the graph has a more understandable range
##Answer: The relationship between newborn body mass and gestation is a positive correlation. The longer the gestation period, the larger the mammal would be at birth.
```


## 7. You should notice that because of the outliers in newborn mass, we need to make some changes. We didn't talk about this in lab, but you can use `scale_x_log10()` as a layer to correct for this issue. This will log transform the y-axis values.
```{r}
life_history %>% 
  ggplot(aes(x=newborn, y=gestation))+
  scale_x_log10()+ ## removes outliers from newborn mass data
  geom_jitter()+
   geom_smooth(method=lm, se=FALSE)
## This is the same function as the code before, but this coded uses scale_x_log10() instead of log10().
```


## Do not do #8,9,10 for Homework 5!
## 8. Now that you have the basic plot, color the points by taxonomic order.

## 9. Lastly, make the size of the points proportional to body mass.

## 10. Make a plot that shows the range of lifespan by order.
